---
title: "Final Project: Exploratory Data Analysis"
format: html
editor: visual
---

## Exploratory Data Analysis (EDA)

### Introduction

#### About the dataset:

The data for this project is from **Diabetes Health Indicators Dataset**. This dataset was put together from the Center of Disease Control's (CDC) Behavior Risk Factor Surveillance System (BRFSS) in 2015. The BRFSS is a premier telephone health-related telephone survey designed to collect state data about United States residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services. We are particularly interested in the indicators for Diabetes.

#### Variables from the dataset that we'll work with:

**Diabetes_binary**: If subject has diabetes or not. While the dataset source mentions a prediabetes level, there are no corresponding values in the dataset.

0 = no diabetes\
1 = diabetes

**HighBP**: If subject has high blood pressure or not.

0 = no high blood pressure\
1 = high blood pressure

**HighChol**: If subject has high cholesterol or not.

0 = no high cholesterol\
1 = high cholesterol

**CholCheck**: If subject has had their cholesterol checked in the past 5 years.

0 = no cholesterol check in 5 years\
1 = yes cholesterol check in 5 years

**BMI**: Body Mass Index.

**PhysActivity**: Physical activity in past 30 days - not including job.

0 = no\
1 = yes

**Sex**: Is the subject male or female.

0 = female\
1 = male

**Age**: Thirteen level category indicating the age of the subject.

1 = 18-24\
2 = 25-29\
3 = 30-34\
4 = 35-39\
5 = 40-44\
6 = 45-49\
7 = 50-54\
8 = 55-59\
9 = 60-64\
10 = 65-69\
11 = 70-74\
12 = 75-79\
13 = 80 or older

#### Purpose of Exploratory Data Analysis:

The purpose of Exploratory Data Analysis (EDA) is to understand the underlying structure, patterns, and relationships in a dataset before applying statistical models or making predictions. EDA serves as a critical first step in data analysis to ensure data quality and to guide further analysis or modeling efforts.

#### Ultimate goal of modeling:

The ultimate goal of modeling in data science and statistics is to create a representation of the data's underlying processes that can be used to 1) Explain Relationships, 2) Make Predictions, 3) Optimize Systems, 4) Classify or Group Data, 5) Identify Key Variable, 6) Reduce Complexity, 7) Generate Insights, and 8) Facilitate Generalization.

### Reading in Diabetes Data Set and Selecting our Chosen Variables

```{r}
#| echo: true
#| include: false
library(readr)
library(ggplot2)
library(tidyverse)
library(forcats)
```

```{r}
#| echo: false
# Reading in CSV file using relative path
data <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

```{r}
# Selecting chosen variables
data <- data |>
  select(Diabetes_binary, HighBP, HighChol, CholCheck, BMI, PhysActivity, Sex, Age)
```

### Exploratory Data Analysis

#### Understanding how the data is stored

```{r}
# Seeing how the variables are stored
str(data)
```

#### Basic Data Validation

```{r}
# Column names
names(data)
```

```{r}
# Looking at unique values in the Diabetes_binary variable
unique(data$Diabetes_binary)
```

```{r}
# Quick numeric summary of data
summary(data)
```

**Every variable except BMI and Age are binary, as indicated CDC's dataset description as well as by the min and max values.**

#### Looking for Missing Values

```{r}
# Using tidy syntax to look for missing values
data |>
  is.na() |>
  colSums()
```

**There are no missing values in this dataset.**

#### Applying Transformations

```{r}
data <- data |>
  mutate(
    # Converting most variables to factors
    diabetes = as.factor(Diabetes_binary),
    high_bp = as.factor(HighBP),
    high_chol = as.factor(HighChol),
    chol_check = as.factor(CholCheck),
    physical_act = as.factor(PhysActivity),
    sex = as.factor(Sex),
    age = as.factor(Age),
    
    # Recoding factor levels to change the names of the values 
    diabetes = fct_recode(diabetes,
                          `No Diabetes` = "0",
                          Diabetes = "1"),
    high_bp = fct_recode(high_bp,
                         No = "0",
                         Yes = "1"),
    high_chol = fct_recode(high_chol,
                           No = "0",
                           Yes = "1"),
    chol_check = fct_recode(chol_check,
                            No = "0",
                            Yes = "1"),
    physical_act = fct_recode(physical_act,
                              No = "0",
                              Yes = "1"),
    sex = fct_recode(sex,
                     Female = "0",
                     Male = "1"),
    age = fct_recode(age,
                     `18-24` = "1",
                     `25-29` = "2",
                     `30-34` = "3",
                     `35-39` = "4",
                     `40-44` = "5",
                     `45-49` = "6",
                     `50-54` = "7",
                     `55-59` = "8",
                     `60-64` = "9",
                     `65-69` = "10",
                     `70-74` = "11",
                     `75-79` = "12",
                     `80_or_older` = "13")) |>
  # Removing most original variables
  select(-Diabetes_binary, -HighBP, -HighChol, -CholCheck, -PhysActivity, -Sex, -Age)
```

#### Investigating Distributions

```{r}
# Investigating the Response Variable
table("diabetes" = data$diabetes)
```

**Many more subjects without diabetes then with diabetes.**

```{r}
# Investigating the Response Variable and Blood Pressure
table("diabetes" = data$diabetes,
      "High Blood Pressure" = data$high_bp)
```

**Roughly 62% of subjects without diabetes reported no high blood pressure and roughly 25% of subjects with diabetes reported no high blood pressure.***

```{r}
# Investigating the Response Variable and Cholesterol
table("diabetes" = data$diabetes,
      "High Cholesterol" = data$high_chol)
```

**There is roughly four times more people with high cholesterol without diabetes then with diabetes.**

```{r}
# Investigating the Response Variable and Sex
table("diabetes" = data$diabetes,
      "Sex" = data$sex)
```

**The difference between males and females with diabetes in this study is roughly 1,500 people. In a dataset consisting of 253,680 people, this is exteremely close.**

```{r}
# Investigating the Response Variable by Sex and Blood Pressure
table("diabetes" = data$diabetes,
      "High Blood Pressure?" = data$high_bp,
      "Sex" = data$sex)
```

**Similar to the last the contingency table, the number of males and females with and without diabetes and high blood pressure is very close.**

```{r}
# Investigating the Response Variable and Age
table("diabetes" = data$diabetes,
      "Age Range" = data$age)
```

**Subjects in the age range of 65-69 made up the largest group of individuals with diabetes.** 

```{r}
# Measure of center and spread for BMI grouped by diabetes
data |>
  group_by(diabetes) |>
  summarize(across("BMI", 
                   .fns = c("mean" = mean,
                            "median" = median,
                            "sd" = sd,
                            "IQR" = IQR,
                            "min" = min,
                            "max" = max),
                   .names = "{.col}_{.fn}"))
```

**The mean and median BMI values for individuals with and without diabetes are virtually the same.**

#### Visualizing the Data

```{r}
# Bar plot
ggplot(data, aes(x = diabetes)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of Individuals with Diabetes",
       x = "Diabetes?",
       y = "Count") +
  theme(plot.title = element_text(hjust = 0.5))
```

**It's interesting to visualize just how many less people there are with diabetes then without in this dataset.**

```{r}
# Stacked bar plot
ggplot(data, aes(x = diabetes, fill = sex)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Sex",
       x = "Diabetes?",
       y = "Proportion") +
  scale_fill_manual(values = c("skyblue", "salmon")) +
  theme(plot.title = element_text(hjust = 0.5))
```

**Again, the difference between males and females with diabetes in this study is roughly 1,500 people. It's interesting that more males have diabetes then females.**

```{r}
# Stacked bar plot
ggplot(data, aes(x = diabetes, fill = physical_act)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Physical Activity in the Last 30 Days",
       x = "Diabetes?",
       y = "Proportion") +
  scale_fill_manual(values = c("skyblue", "salmon")) +
  theme(plot.title = element_text(hjust = 0.5))
```

**It's interesting, and unsurprising, that individuals with diabetes engage in less physical activity.** 

```{r}
ggplot(data, aes(x = diabetes, y = BMI)) +
  geom_jitter(aes(color = sex)) +
  labs(title = "Diabetes and Body Mass Index by Sex",
       x = "Diabetes?",
       y = "BMI") +
  theme(plot.title = element_text(hjust = 0.5))
```

**The scatter plot clearly shows that individuals with diabetes tend to have a lower BMI.**

```{r}
# Proportional bar plot for Diabetes by Physical Activity and Age
ggplot(data, aes(x = diabetes, fill = physical_act)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Physical Activity and Age",
       x = "Diabetes?",
       y = "Proportion") +
  scale_fill_manual(values = c("skyblue", "salmon")) +
  facet_wrap(~age) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 60, hjust = 1))  # Rotate x-axis text
```

**It's interesting that at the lowest age range (18-24) the proportion of individuals with and without diabetes is the closest.**

```{r}
# Proportional bar plot for Diabetes by Sex and Blood pressure
ggplot(data, aes(x = diabetes, fill = high_bp)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Diabetes by Physical Activity and Age",
       x = "Diabetes?",
       y = "Proportion") +
  scale_fill_manual(values = c("skyblue", "salmon")) +
  facet_wrap(~sex) +
  theme(plot.title = element_text(hjust = 0.5))
```

**It's interesting that the proportions of males and females with and without diabetes are similar.**